Composite Bayesian inference

### 2019-01-15 papers/composite_inference/draft.tex
Modifs draft CL

\begin{abstract}
This note revisits the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents. Our view makes it natural to use a machine learning approach to calibrate composite likelihood, that is, tune the associated weights. For instance, the weights can be set to minimize cross-entropy on a labeled training dataset, as is a typical supervised learning strategy, yielding an easy-to-solve convex problem. We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to some artificial intelligence tasks.
\end{abstract}

@@

Classical frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large training sets, including ``shallow'' learning techniques (maximum entropy classifiers \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} with the exception of deep belief networks \cite{Hinton-06,Fischer-14}.

A common criticism of discriminative models is that they are ``black boxes'' in that they can provide a valuable answer to an inference problem, but no explanation to support it. This is problematic in application fields such as medical diagnosis, where AI is intended as a tool to assist humans in a decision making process, but decision is not delegated to AI. Automated disease predictions should therefore be corroborated by empirical evidence -- in other words, AI systems should be able to justify their opinion just like human experts. Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide useful clues to clinicians despite being less powerful in terms of prediction performance.

@@

Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters unless the variable of interest is observed. This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models. For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence it is less effective in small training sets \cite{Ng-01}.

@@

Composite likelihood (CL, see \cite{Varin-11} and the references therein) is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model. The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole. This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood. While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.

In this note, CL is interpreted as a probabilistic log-linear opinion pool of ``agents'' making use of different pieces of information extracted from the input data. Each agent acts as an isolated statistician using a generative model of a single feature to express an opinion in the form of a likelihood function of the hidden variables. The agent opinions are then aggregated into a predictive probability distribution analogous to a Bayesian posterior.

The relevance of each feature in the opinion pool is determined by an arbitrary weight used to scale the associated likelihood. For classification performance, the weights can be tuned in a typical supervised learning scenario so as to concurrently maximize discriminative likelihood (or, equivalently, minimize cross-entropy), thereby making the method hybrid between generative and discriminative modeling approaches. We shall note that this particular training strategy amounts to building a maximum entropy classifier \cite{BergerA-96} that uses the feature-based log-likelihoods as basis functions.

{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression. Double training phase.}

{\color{red} Interpretability of composite likelihood. Obvious from the construction. Can achieve prediction performance comparable with fully discriminative models while maintaining the interpretability of generative models.}

@@

\section{Composite likelihood as opinion pooling}
\label{sec:pool}

@@

This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated. A natural workaround known as {\em data reduction} is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(x) = p(z|x)
.
$$

@@

If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. We will show in Section~\ref{sec:maxent} that one that is particularly well-suited to our case is the {\em log-linear opinion pool}: 
\begin{equation}
\label{eq:log_pool}
p_\lambda(x|y) = \frac{1}{z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is some reference distribution or prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights so that the normalizing factor:
$$
z_\lambda(y) = \int \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i} dx
$$
is finite, which holds whenever $\lambda\succeq 0$, {\em i.e.} all weights are positive, assuming that the agent opinions are always upper bounded. Negative weights should be ruled out as the existence condition $z_\lambda(y)<\infty$ would then depend on the particular opinions expressed by the agents. Also note that, while it is possible to have the weights depend on~$y$, there is a strong rational for constant weights as we shall see in the sequel.

@@

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding the form: $p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x)$, where the quantity:
\begin{equation}
\label{eq:comp_lik}
L^c_\lambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays the same role as a traditional likelihood function. This expression happens to be known in statistics as a {\em marginal composite likelihood} \cite{Varin-11}. See Appendix~\ref{sec:conditional} for the slightly more general form called {\em conditional composite likelihood}, which can be derived in the same way.

From~(\ref{eq:comp_lik}), we see that CL shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$. The clear computational advantage of CL over the true likelihood is that it only requires to evaluate the marginal feature distributions rather than the joint distribution of all features.

\section{Tuning composite likelihood weights}

When the features can be considered exchangeable, it is natural to choose uniform CL weights. The CL is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE). The weight may be tuned so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match the frequentist and composite Bayesian notions of uncertainty.

Ignoring such a goal, another frequent recommandation for CL weights is to sum up to one. This can be motivated in various ways. It turns out that the only pooling operator that does not explicitly depend on~$x$ and preserves {\em external Bayesianity} is the log-linear opinion pool with unit sum weights \cite{Genest-86b}. External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior. Another appealing property of log-linear pooling with unit sum weights is to minimize the average Kullback-Leibler (KL) divergence to the agent opinions \cite{Garg-04}. A maximum relative entropy property is also given in \cite{Wang-14}, Theorem~3.

Unit sum weights, however, correspond to the extreme situation where the features are assumed to be maximally redundant, but is clearly ineffective for independent or weakly correlated features, which require weights close to one for the CL to closely approximate the true likelihood. In this case, the CL may be much flatter than it should, leading to severly overestimated credibility sets.

{\color{red} Why is it important to fine-tune the weights?}
We therefore need a method to tune weights that can automatically adjust to the level of statistical dependence beween features.

@@

\section{Maxent composite likelihood}
\label{sec:maxent}

What does the constraint mean? It means that a model is admissible if it compresses the data at least as well as all my agents. So we are not assuming that the generative models are {\em true}. We only assume that they are useful to compress the (reduced) data. Connection to MDL.

@@

>>>>

\begin{abstract}
This note revisits the concept of composite likelihood as a method to make a probabilistic inference by aggregation of multiple Bayesian agents. This view makes it natural to use a machine learning approach to calibrate the weights associated with composite likelihood. For instance, they can be tuned in a typical supervised learning phase so as to minimize cross-entropy on a labeled training dataset, yielding a convex problem. We argue that composite Bayesian inference is a middle way between generative and discriminative modeling approaches that trades off between interpretability and prediction performance, both of which are crucial to some artificial intelligence tasks.
\end{abstract}

@@

Frequentist and Bayesian inference paradigms rest upon the existence of a probabilistic data-generating model that is both empirically valid and computationally tractable. Because this is challenging for complex data, other inference models are commonly used in applied science: deliberately misspecified generative models, as in quasi-likelihood methods \cite{White-82,Walker-13} or na\"ive Bayes \cite{Ng-01}; data compression models as in minimum description length \cite{Grunwald-07}; and discriminative models\footnote{A {\em discriminative} model is a parametric family that describes the distribution of a variable of interest conditional on data, in contrast with a {\em generative} model, which describes the conditional distribution of data given the variable of interest.}, which currently dominate artificial intelligence (AI) and typically require supervised learning on large datasets, including ``shallow'' learning techniques (maximum entropy classifiers aka logistic regression \cite{BergerA-96}, support vector machines \cite{Vapnik-00}, Gaussian processes \cite{Rasmussen-06}) as well as most deep learning techniques \cite{Lecun-15,Goodfellow-16} (with the exception of deep belief networks \cite{Hinton-06,Fischer-14}).

A common criticism of discriminative models is that they are ``black boxes'' in that they can make valid inferences but provide little or no information to support them \cite{Molnar-18}. This is problematic in application fields such as medical diagnosis, where correct inferences are not straightforward to humans and AI is intended as a tool to assist a decision making process. Automated disease predictions should therefore be corroborated by individualized findings to be of any clinical value -- in other words, AI systems should be able to justify their opinion just like clinicians. Unlike discriminative models, simple methods based on range analysis of univariate biomarkers can provide meaningful clues, and are thus more useful in practice although less powerful in terms of prediction performance.

@@

Another limitation of discriminative models is that they are not suitable for unsupervised learning or on-the-fly parameter estimation because they treat the data and the model parameters as marginally independent, meaning that the data conveys no information about the parameters unless the variable of interest is observed. This is illustrated in Figure~\ref{fig:graph_comparison} by the respective directed graph representations of generative and discriminative models. For the same reason, supervised learning in a discriminative model is less precise than in a generative model spanning the same family of posteriors, hence is less effective in small training sets \cite{Ng-01}.

@@

This note advocates the (old) idea of probabilistic opinion pooling \cite{Genest-86} as a possible compromise between generative and discriminative models. Rather than using a purely discriminative model or attempting at a full generative model, we may combine multiple low-dimensional generative models for different pieces of information extracted from the input data. Each such model acts as an isolated ``agent'' that uses a single feature to express an opinion in the form of a likelihood function of the variable of interest. The agent opinions may then be aggregated into a unique predictive probability distribution analogous to a Bayesian posterior. This simple idea may be understood as a probabilistic version of boosting, or as a ``divide and conquer'' approximation to the intractable Bayesian posterior. Importantly, predictions made in such a way could be readily interpretable owing to the use of low-dimensional generative models.

It turns out that, when the aggregation method is chosen as a log-linear pool, the consensus distribution is (up to a multiplicative prior) a quantity known as composite likelihood (CL); see \cite{Varin-11} and the references therein. CL was developed in computational statistics as an extension of the familiar notion of likelihood to situations where the likelihood is intractable. While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is proven to be consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.

%the data distribution is not modeled as a whole but separate feature-based generative models are available

%CL is a semi-generative approach to statistical inference that extends the familiar notion of likelihood without requiring a full generative model. The key idea is to model an arbitrary set of low-dimensional features separately and then combine them, instead of modeling the data distribution as a whole. This may be viewed as a ``divide and conquer'' method to approximate the true but intractable likelihood. While maximum CL does not inherit the general property of maximum likelihood to yield an asymptotically minimum-variance estimator, it is consistent under mild conditions \cite{Xu-11} and may offer an excellent trade-off between computational and statistical efficiency in practice.

%An open question regarding CL is how to choose the weights associated with the respective features.

With the opinion pooling interpretation in mind, the feature weights in CL may be optimized for prediction performance in a typical discriminative learning scenario. As we will see, this strategy amounts to training a maximum entropy classifier using the feature-based log-likelihoods as basis functions. These likelihood functions themselves may need to be pre-trained, therefore the proposed model typically comes with a double training scheme: a generative training phase (to learn feature-based likelihood parameters) followed by a discriminative training phase (to learn feature weights in aggregation).

Technical details are given in the remainder.

%The relevance of each feature in the CL is determined by an arbitrary weight used to scale the associated likelihood. For classification performance, the weights can be tuned in a typical supervised learning scenario so as to concurrently maximize discriminative likelihood (or, equivalently, minimize cross-entropy), thereby making the method hybrid between generative and discriminative modeling approaches. We shall note that this particular training strategy amounts to building a maximum entropy classifier \cite{BergerA-96} that uses the feature-based log-likelihoods as basis functions.

%{\color{red} Still, note that the likelihood functions themselves are trained beforehand so this is slightly different from a standard maxent classifier aka logistic regression. Double training phase.}

%{\color{red} Classical composite likelihood not studied with nuisance parameters. In our scenario, nuisance parameters are learned beforehand.}

%{\color{red} Interpretability of composite likelihood. Obvious from the construction. Can achieve prediction performance comparable with fully discriminative models while maintaining the interpretability of generative models.}

@@

\section{Composite likelihood as opinion pooling}
\label{sec:log_pool}

@@

This requires a plausible generative model which, for complex data, may be out of reach or involve too many parameters to be estimated. A natural workaround known as data reduction is to extract some lower-dimensional representation $z=f(y)$, where $f$ is a many-to-one mapping, and consider the potentially more convenient likelihood function:
$$
\ell(x) = p(z|x)
.
$$

@@

If we see the likelihoods as posterior distributions corresponding to uniform priors, this is a classical problem of probabilistic opinion aggregation from possibly redundant sources, for which several methods exist in the literature \cite{Tarantola-82,Genest-86,Garg-04,Allard-12}. One that is appealing as it tends to produce single peaked distributions is the {\em generalized logarithmic opinion pool}:
\begin{equation}
\label{eq:log_pool}
p_\lambda(x|y) = \frac{1}{z_\lambda(y)} \pi(x) \prod_{i=1}^n p(z_i|x)^{\lambda_i},
\end{equation} 
where $\pi(x)$ is some positive-value function which can be interpreted as a global prior, and $\lambda=(\lambda_1,\ldots,\lambda_n)$ is a vector of weights chosen so that the normalizing factor $z_\lambda(y)$ is finite. This condition holds whenever all weights are positive ($\lambda\succeq 0$), assuming that the agent opinions are always upper bounded. Negative weights should be ruled out to warrant a natural monotonicity property: if all agents agree that an outcome $x_1$ is less likely than an outcome $x_2$, then the consensus probabilities should always satisfy $p_\lambda(x_1|y)\leq p_\lambda(x_2|y)$. Note that it is possible to have the weights depend on the data~$y$, but this would make the method considerably more complex and will therefore not be considered here.

@@

\section{Composite Bayes rule}
\label{sec:bayes_rule}

The log-linear pool~(\ref{eq:log_pool}) bears a striking similarity to Bayes rule, yielding  the form: 
$$
p_\lambda(x|y)\propto \pi(x) L^c_\lambda(x),
$$
where the function:
\begin{equation}
\label{eq:comp_lik}
L^c_\lambda(x) \equiv \prod_{i=1}^n \ell_i (x)^{\lambda_i}
\end{equation} 
plays the same role as a traditional likelihood function. This expression happens to be known as a {\em marginal composite likelihood} \cite{Varin-11}. The slightly more general {\em conditional composite likelihood} form can be derived in the same way as above by conditioning all probabilities on confounding variables, see Appendix~\ref{sec:conditional}. The computational advantage of CL over a genuine likelihood function stems from the fact that it is more efficient to evaluate the marginal distributions of each feature  than the joint distribution of all features.

CL in equation~(\ref{eq:comp_lik}) shares a convenient factorized form with the likelihood derived under the assumption of mutual feature independence, usually referred to as {\em na\"ive Bayes} or {\em simple Bayes} in the machine learning literature, which corresponds to the special case of unitary weights, $\lambda_1=\ldots=\lambda_n= 1$. While our derivation does not assume feature independence, it yields an alternative interpretation of na\"ive Bayes as a general feature aggregation method. However, unitary CL weights do not guarantee optimal prediction performance.

\section{Composite weight tuning}

\subsection{A priori tuning}

%When the features can be considered exchangeable, it is natural to choose constant CL weights.

In the absence of knowledge, a default rule is to choose constant CL weights. The CL function is then a scaled version of the na\"ive Bayes likelihood, the common weight value being irrelevant to the maximum CL estimator (MCLE). For the purpose of computing reasonable credibility sets, it may be tuned empirically so as to best adjust the pseudo posterior variance matrix to the asymptotic variance matrix of the MCLE \cite{Pauli-11}, or via a close-in-spirit curvature adjustment \cite{Ribatet-12}, in attempts to match frequentist and composite Bayesian notions of uncertainty.

Ignoring such a goal, another sensible prior recommendation for log-linear pooling weights is to sum up to one. This is motivated in \cite{Genest-86b} by the fact that the only pooling operator that preserves external Bayesianity\footnote{External Bayesianity essentially means that it should not matter whether the prior is incorporated before or after pooling, provided that all agents agree on the same prior.} is the log-linear opinion pool with unit sum weights. A further justification given in \cite{Garg-04} is that the log-linear pool with unit sum weights minimizes the average Kullback-Leibler (KL) divergence to the agent opinions. See also \cite{Wang-14}, Theorem~3, for another information theoretic argument.

Unit sum weights, however, implicitly assume strong redundancy between features and may therefore be ineffective for moderately correlated features. An ideal method to tune weights is one that can capture and adjust for existing statistical dependences in the data. This calls for a training approach, which, as we will see, is unlikely to pick constant or unit sum weights.

%automatically adjust to the level of statistical dependence beween features.

%are over-conservative for , which require weights close to one for the CL to closely approximate the true likelihood. Using weights smaller than one, the CL will be flatter than it should and lead to severly overestimated credibility sets.

%correspond to the extreme situation where the features are assumed to be maximally redundant, but is

%{\color{red} Why is it important to fine-tune the weights?}

\subsection{Training}

Brute force approach: optimize fit to Bayesian posterior.

@@

\section{Extensions}

@@

\section{Maxent}

\subsection{Training}
\label{sec:maxent}

{\color{red}This stuff should move to the appendix. It is not critical to understand the Maxent connection of trained composite likelihood, althought it gives some insight. Essentially, agents report compressing rates using feature-based models and we either trust them (equality constraints) or assume that they are underestimated (inequality constraints). Why underestimated? Well, it's just more general. Overestimated leads nowhere, we have to assume something.}

What does the inequality constraint mean? It means that a model is admissible if it jointly allows for all agents to achieve the same compression rate {\em or better} than achieved individually. Somewhow, we forget that there should exist a true distribution that is consistent with the agent constraints both jointly and individually. We forget that not because it is not true but because we deem it irrelevant. Connection to MDL? We are reminded of these words by Rissanen:
\begin{quote}
No statistical model is ``true" or ``false", ``right" or ``wrong"; the models just have varying performance, which can be assessed.  
\end{quote}

Importantly, no agent will be proven wrong since individual performance is at least as good as announced. But if it is strictly better, it means that the agent is simply discarded (getting a zero weight rather than a difficult-to-intepret negative weight).

We could also consider inequality constraints as a relaxation: explicitly making it easier for Nature to screw us up, but for a good cause. Not to make optimization easier, to make it easier to intepret. Why? Because we know that a posteriori from the fact that we get $\lambda_i\geq 0$. But a priori?

%Okay, but then it's not ``nature'' playing, is it? The game is between a ``generator'' G and a ``predictor'' P. G tries to make it as difficult as possible to predict outcomes using an admissible model. We could argue that true models are at the boundaries of the admissible space, but G is not restricted to true models as long as it maintains the agents performance level.

%In fact, P is the agent coordinator and G could be seen as some sort of test controller. Inequality constraints add some further robustness to the model.
